---
title: "Comparing loadings update algorithms"
author: "Jason Willwerscheid"
date: "7/19/2018"
output: 
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

Here I implement the algorithm described in a [previous note](flash_em.html) and compare results with FLASH.

## Code

Click "Code" to view the implementation.

```{r code}

# INITIALIZATION FUNCTIONS ------------------------------------------

fl_to_altfl <- function(data, fl, k) {
  altfl <- list()
  altfl$tau <- fl$tau
  altfl$Rk <- flashr:::flash_get_Rk(data, fl, k)
  altfl$R2k <- flashr:::flash_get_R2k(data, fl, k)
  
  altfl$al <- fl$gl[[k]]$a
  altfl$pi0l <- fl$gl[[k]]$pi0
  altfl$af <- fl$gf[[k]]$a
  altfl$pi0f <- fl$gf[[k]]$pi0

  s2 = 1/(fl$EF2[, k] %*% t(fl$tau))
  s = sqrt(s2)
  Rk = flashr:::flash_get_Rk(data, fl, k)
  x = fl$EF[, k] %*% t(Rk * fl$tau) * s2
  w = 1 - fl$gl[[k]]$pi0
  a = fl$gl[[k]]$a
  
  altfl$wl <- ebnm:::wpost_normal(x, s, w, a)
  altfl$mul <- ebnm:::pmean_cond_normal(x, s, a)
  altfl$s2l <- ebnm:::pvar_cond_normal(s, a)
  
  s2 = 1/(fl$EL2[, k] %*% fl$tau)
  s = sqrt(s2)
  Rk = flashr:::flash_get_Rk(data, fl, k)
  x = fl$EL[, k] %*% (Rk * fl$tau) * s2
  w = 1 - fl$gf[[k]]$pi0
  a = fl$gf[[k]]$a
  
  altfl$wf <- ebnm:::wpost_normal(x, s, w, a)
  altfl$muf <- ebnm:::pmean_cond_normal(x, s, a)
  altfl$s2f <- ebnm:::pvar_cond_normal(s, a)
  
  altfl$KL <- sum(unlist(fl$KL_l)[-k] + unlist(fl$KL_f)[-k])
  
  return(altfl)
}

altfl_to_fl <- function(altfl, fl, k) {
  fl$EL[, k] <- compute_EX(altfl$wl, altfl$mul)
  fl$EL2[, k] <- compute_EX2(altfl$wl, altfl$mul, altfl$s2l)
  fl$EF[, k] <- compute_EX(altfl$wf, altfl$muf)
  fl$EF2[, k] <- compute_EX2(altfl$wf, altfl$muf, altfl$s2f)
  
  fl$gl[[k]] <- list(pi0 = altfl$pi0l, a = altfl$al)
  fl$gf[[k]] <- list(pi0 = altfl$pi0f, a = altfl$af)
  fl$ebnm_fn_l <- fl$ebnm_fn_f <- "alt"
  fl$ebnm_param_l <- fl$ebnm_param_f <- list()
  
  fl$tau <- altfl$tau
  
  return(fl)
}

# OBJECTIVE FUNCTION ------------------------------------------------

compute_obj <- function(altfl) {
  with(altfl, {
    EL <- compute_EX(wl, mul)
    EL2 <- compute_EX2(wl, mul, s2l)
    EF <- compute_EX(wf, muf)
    EF2 <- compute_EX2(wf, muf, s2f)

    obj <- rep(0, 4)
    
    obj[1] <- sum(0.5 * log(tau / (2 * pi)))
    obj[2] <- sum(-0.5 * (tau * (R2k - 2 * Rk * outer(EL, EF) + outer(EL2, EF2))))
    
    obj[3] <- compute_KL(wl, mul, s2l, pi0l, al)
    obj[4] <- compute_KL(wf, muf, s2f, pi0f, af)
  
    return(sum(obj) + KL)
  })
}

compute_KL <- function(w, mu, s2, pi0, a) {
  obj <- rep(0, 3)
  
  tmp <- (1 - w) * (log(pi0) - log(1 - w)) 
  obj[1] <- sum(tmp[!is.nan(tmp)])
  tmp <- w * (log(1 - pi0) - log(w))
  obj[2] <- sum(tmp[!is.nan(tmp)])
  obj[3] <- sum(0.5 * w * (log(a) + log(s2) + 1 - a * (mu^2 + s2)))
  
  return(sum(obj))
}

compute_EX <- function(w, mu) {
  return(as.vector(w * mu))
}

compute_EX2 <- function(w, mu, sigma2) {
  return(as.vector(w * (mu^2 + sigma2)))
}

# UPDATE FUNCTIONS --------------------------------------------------

update_a <- function(w, EX2) {
  return(sum(w) / sum(EX2))
}

update_pi0 <- function(w) {
  return(sum(1 - w) / length(w))
}

update_mul <- function(a, tau, Rk, EF, EF2) {
  n <- nrow(tau)
  p <- ncol(tau)
  numer <- rowSums(tau * Rk * matrix(EF, nrow=n, ncol=p, byrow=TRUE))
  denom <- a + rowSums(tau * matrix(EF2, nrow=n, ncol=p, byrow=TRUE))
  return(numer / denom)
}

update_muf <- function(a, tau, Rk, EL, EL2) {
  n <- nrow(tau)
  p <- ncol(tau)
  numer <- colSums(tau * Rk * matrix(EL, nrow=n, ncol=p, byrow=FALSE))
  denom <- a + colSums(tau * matrix(EL2, nrow=n, ncol=p, byrow=FALSE))
  return(numer / denom)
}

update_s2l <- function(a, tau, EF2) {
  n <- nrow(tau)
  p <- ncol(tau)
  return(1 / (a + rowSums(tau * matrix(EF2, nrow=n, ncol=p, byrow=TRUE))))
}

update_s2f <- function(a, tau, EL2) {
  n <- nrow(tau)
  p <- ncol(tau)
  return(1 / (a + colSums(tau * matrix(EL2, nrow=n, ncol=p, byrow=FALSE))))
}

update_wl <- function(a, pi0, mu, sigma2, tau, Rk, EF, EF2) {
  C1 <- log(1 - pi0) - log(pi0)
  C2 <- 0.5 * (log(a) + log(sigma2) - a * (mu^2 + sigma2) + 1)
  C3 <- rowSums(tau * (Rk * outer(mu, EF) - 0.5 * outer(mu^2 + sigma2, EF2)))
  C <- C1 + C2 + C3
  return(1 / (1 + exp(-C)))
}

update_wf <- function(a, pi0, mu, sigma2, tau, Rk, EL, EL2) {
  C1 <- log(1 - pi0) - log(pi0)
  C2 <- 0.5 * (log(a) + log(sigma2) - a * (mu^2 + sigma2) + 1)
  C3 <- colSums(tau * (Rk * outer(EL, mu) - 0.5 * outer(EL2, mu^2 + sigma2)))
  C <- C1 + C2 + C3
  return(1 / (1 + exp(-C)))
}

# ALGORITHM ---------------------------------------------------------

update_tau <- function(altfl) {
  within(altfl, {
    EL <- compute_EX(wl, mul)
    EL2 <- compute_EX2(wl, mul, s2l)
    EF <- compute_EX(wf, muf)
    EF2 <- compute_EX2(wf, muf, s2f)
    
    R2 <- R2k - 2 * Rk * outer(EL, EF) + outer(EL2, EF2)
    tau <- matrix(1 / colMeans(R2), nrow=nrow(tau), ncol=ncol(tau),
                  byrow=TRUE)
  })
}

update_loadings_post <- function(altfl) {
  within(altfl, {
    EF <- compute_EX(wf, muf)
    EF2 <- compute_EX2(wf, muf, s2f)
    
    mul <- update_mul(al, tau, Rk, EF, EF2)
    s2l <- update_s2l(al, tau, EF2)
    wl <- update_wl(al, pi0l, mul, s2l, tau, Rk, EF, EF2)
  })
}

update_loadings_prior <- function(altfl) {
  within(altfl, {
    EL2 <- compute_EX2(wl, mul, s2l)
    
    al <- update_a(wl, EL2)
    pi0l <- update_pi0(wl)
  })
}
  
update_factor_post <- function(altfl) {
  within(altfl, {
    EL <- compute_EX(wl, mul)
    EL2 <- compute_EX2(wl, mul, s2l)
    
    muf <- update_muf(af, tau, Rk, EL, EL2)
    s2f <- update_s2f(af, tau, EL2)
    wf <- update_wf(af, pi0f, muf, s2f, tau, Rk, EL, EL2)
  })
}

update_factor_prior <- function(altfl) {
  within(altfl, {
    EF2 <- compute_EX2(wf, muf, s2f)
    
    af <- update_a(wf, EF2)
    pi0f <- update_pi0(wf)
  })
}

do_one_update <- function(altfl) {
  obj <- rep(0, 5)
  
  altfl <- update_tau(altfl)
  obj[1] <- compute_obj(altfl)
  
  altfl <- update_loadings_post(altfl)
  obj[2] <- compute_obj(altfl)
  
  altfl <- update_loadings_prior(altfl)
  obj[3] <- compute_obj(altfl)
  
  altfl <- update_factor_post(altfl)
  obj[4] <- compute_obj(altfl)
  
  altfl <- update_factor_prior(altfl)
  obj[5] <- compute_obj(altfl)
  
  return(list(altfl = altfl, obj = obj))
}

optimize_alt_fl <- function(altfl, tol = .01, verbose = FALSE) {
  obj <- compute_obj(altfl)
  diff <- Inf
  
  while (diff > tol) {
    tmp <- do_one_update(altfl)
    new_obj <- tmp$obj[length(tmp$obj)]
    diff <- new_obj - obj
    obj <- new_obj
    if (verbose) {
      message(paste("Objective:", obj))
    }
    altfl <- tmp$altfl
  }
  
  return(altfl)
}
```

## Fit

Using the same dataset as in previous investigations, I fit a FLASH object with four factors (recall that it's the fourth factor that was causing problems during loadings updates):

```{r flfit}
load("./data/before_bad.Rdata")

# devtools::install_github("stephenslab/flashr")
devtools::load_all("/Users/willwerscheid/GitHub/flashr")
# devtools::install_github("stephenslab/flashr")
devtools::load_all("/Users/willwerscheid/GitHub/ebnm")

fl <- flash_add_greedy(data, Kmax=4, verbose=FALSE)
```

The objective as computed by FLASH is:

```{r flobj}
flash_get_objective(data, fl)
```

I now convert the fourth factor to an "altfl" object. The objective as computed by the alternate method is:
```{r altfl}
altfl <- fl_to_altfl(data, fl, 4)
compute_obj(altfl)
```

So the objective functions agree. This is a good thing.

Next, I attempt to optimize the altfl object:
```{r opt_altfl}
altfl <- optimize_alt_fl(altfl, verbose=TRUE)
```

Nothing happens. This is also a good thing!

## Conclusion

That I get the same values in both cases confirms that a) the expressions in the [previous note](flash_em.html) are correct; b) the `KL_l` and `KL_f` values obtained using FLASH are in fact reliable.

The algorithm implemented here is, however, very prone to getting stuck in local maxima, and does not appear to be a suitable substitute for the existing FLASH algorithm.
