---
title: "Tensor Decomposition of GTEx Brain Data"
author: "Jason Willwerscheid"
date: "1/8/2019"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Data

TODO: describe data here.

## Code

for preprocessing GTEx data and fitting flash objects to the brain subtensor...
```{r preprocess, code=readLines("../code/brain.R"), eval=FALSE}
```

...and for producing the plots and tables below.
```{r allCode}
# Flip signs so that individual loadings are mostly postive -------------------

align.data <- function(brain) {
  for (k in 1:brain$n.factors) {
    if (mean(brain$loadings$normalized.loadings[[2]][, k]) < 0) {
      brain$loadings$normalized.loadings[[2]][, k] <- -1 *
        brain$loadings$normalized.loadings[[2]][, k]
      brain$loadings$normalized.loadings[[3]][, k] <- -1 *
        brain$loadings$normalized.loadings[[3]][, k]
    }
  }
  return(brain)
}

# Barplots --------------------------------------------------------------------

do.plots <- function(brain, k, incl.gene.plot = TRUE, fix.ind.ylim = FALSE) {
  brain.colors <- c("hotpink", "gray40", "green3", "blue1", "blue3", 
                    "gray20", "black", "lightpink", "red1", "green4", 
                    "green1", "red4", "red3")

  tissue.idx <- order(brain$loadings$normalized.loadings[[1]][, k],
               decreasing = TRUE)
  barplot(brain$loadings$normalized.loadings[[1]][tissue.idx, k],
          col = brain.colors[tissue.idx], 
          axisnames = FALSE,
          xlab = "Tissues")
  
  if (incl.gene.plot) {
    barplot(sort(brain$loadings$normalized.loadings[[3]][, k],
                 decreasing = TRUE),
            axisnames = FALSE,
            xlab = "Genes",
            main = paste0("Factor ", k, ": ", 
                          100 * round(brain$pve[k], 4), "% PVE"))
  }
  
  vals <- brain$loadings$normalized.loadings[[2]][, k]
  exclusions <- brain$fit$exclusions[[k]][[2]]
  if (length(exclusions) > 0)
    vals <- vals[-exclusions]
  ylim <- NULL
  if (fix.ind.ylim)
    ylim <- c(-0.15, 0.25)
  barplot(sort(vals, decreasing = TRUE),
          axisnames = FALSE,
          xlab = "Individuals",
          ylim = ylim)
}

all.plots <- function(brain) {
  par(mfrow = c(2, 3))
  for (k in order(brain$pve, decreasing = TRUE)) {
    do.plots(brain, k)
  }
}

all.plots.comparison <- function(brain1, brain2) {
  par(mfrow = c(2, 4))
  for (k in 1:brain1$n.factors) {
    do.plots(brain1, k, incl.gene.plot = FALSE, fix.ind.ylim = TRUE)
    do.plots(brain2, k, incl.gene.plot = FALSE, fix.ind.ylim = TRUE)
  }
}

# Top tissues and genes -------------------------------------------------------

get.top.tissues <- function(brain, k) {
  tissue.names <- c("amygdala", "ac cortex", "caudate bg",
                    "cerebral hemisphere", "cerebellum",
                    "cortex", "fr cortex", "hippocampus",
                    "hypothalamus", "nuc acc bg", "putamen bg",
                    "spinal cord", "substantia nigra")
  
  loadings <- round(brain$loadings$normalized.loadings[[1]][, k], 2)
  tissue.labels <- paste0(tissue.names, " (", loadings, ")")
  
  n.tissues <- sum(loadings > 0.1)
  if (n.tissues > 6)
    return("most tissues")

  tissue.idx <- order(loadings, decreasing = TRUE)[1:n.tissues]
  return(paste(tissue.labels[tissue.idx], collapse = ", "))
}

get.top.go <- function(brain, k, n.loadings = 200) {
  loadings <- brain$loadings$normalized.loadings[[3]][, k]
  names(loadings) <- sapply(strsplit(names(loadings), ".", fixed = TRUE), 
                            `[`, 1)
  overexpressed <- names(sort(loadings, decreasing = TRUE)[1:n.loadings])
  cp.res <- clusterProfiler::enrichGO(overexpressed, "org.Hs.eg.db", 
                                      ont = "BP", keyType = "ENSEMBL",
                                      universe = names(loadings), 
                                      minGSSize = 10)
  over <- paste(head(cp.res, 3)$Description, collapse = ", ")
  
  if (k == 1) {
    under <- NA
  } else {
    underexpressed <- names(sort(loadings, decreasing = FALSE)[1:n.loadings])
    cp.res2 <- clusterProfiler::enrichGO(underexpressed, "org.Hs.eg.db", 
                                         ont = "BP", keyType = "ENSEMBL",
                                         universe = names(loadings), 
                                         minGSSize = 10)
    under <- paste(head(cp.res2, 3)$Description, collapse = ", ")
  }
  
  list(over = over, under = under)
}

# Regression on demographic and technical variables ---------------------------

do.regression <- function(brain, covar, k) {
  excl <- brain$fit$exclusions[[k]][[2]]
  loadings <- brain$loadings$normalized.loadings[[2]][, k]
  if (length(excl) > 0) {
    covar <- covar[-excl, ]
    loadings <- loadings[-excl]
  }
  df <- cbind(covar, loadings = loadings)
  mod <- lm(loadings ~ ., data = df)
  
  p.vals <- round(coefficients(summary(mod))[-1, 4], 3)
  sex.col <- which(substr(names(p.vals), 1, 1) == "S")
  age.col <- which(substr(names(p.vals), 1, 1) == "A")
  race.col <- which(substr(names(p.vals), 1, 1) == "R")
  tech.col <- (1:length(p.vals))[-c(sex.col, age.col, race.col)]
  
  mod.anova <- anova(mod)
  resid.ss <- mod.anova$`Sum Sq`[8]
  sex.pve <- mod.anova$`Sum Sq`[1] / resid.ss
  age.pve <- mod.anova$`Sum Sq`[2] / resid.ss
  race.pve <- mod.anova$`Sum Sq`[3] / resid.ss
  tech.pve <- sum(mod.anova$`Sum Sq`[4:7]) / resid.ss
  
  sex.res <- paste0("PVE: ", 100 * round(sex.pve, 4), "% (p: ",
                    paste(p.vals[sex.col], collapse = ", "), ")")
  age.res <- paste0("PVE: ", 100 * round(age.pve, 4), "% (p: ",
                    paste(p.vals[age.col], collapse = ", "), ")")
  race.res <- paste0("PVE: ", 100 * round(race.pve, 4), "% (p: ",
                    paste(p.vals[race.col], collapse = ", "), ")")
  tech.res <- paste0("PVE: ", 100 * round(tech.pve, 4), "% (p: ",
                    paste(p.vals[tech.col], collapse = ", "), ")")
  
  return(list(sex = sex.res, age = age.res, race = race.res, tech = tech.res))
}

# Table of results ------------------------------------------------------------

do.table <- function(brain, covar, k) {
  top.go <- get.top.go(brain, k)
  regress.res <- do.regression(brain, covar, k)
  tabl <- rbind(c("PVE: ", paste0(100 * round(brain$pve[k], 4), "%")),
                c("top tissues: ", get.top.tissues(brain, k)),
                c("overexpressed: ", top.go$over),
                c("underexpressed: ", top.go$under),
                c("age effect: ", regress.res$age),
                c("sex effect: ", regress.res$sex),
                c("race effect: ", regress.res$race),
                c("technical factors:", regress.res$tech))
  return(tabl)
}

all.tables <- function(brain, covar) {
  cat("\n")
  for (k in order(brain$pve, decreasing = TRUE)) {
    print(knitr::kable(do.table(brain, covar, k), 
                       caption = paste("Factor", k)))
    cat("\n")
  }
}
```

## A first attempt

At first, results look good. Different factors clearly correspond to different types of brain tissue. But there are problems....

```{r firstAttempt}
brain00 <- readRDS("./data/brain/brain00.rds")
brain00 <- align.data(brain00)
all.plots(brain00)
```

## Using a nonmissingness threshold

Results look better.

```{r brain05}
brain05 <- readRDS("./data/brain/brain05.rds")
brain05 <- align.data(brain05)

brain001 <- readRDS("./data/brain/brain001.rds")
brain001 <- align.data(brain001)

all.plots(brain05)
```

## Factor details

Note, in particular, the large differences from the paper in proportions of variance explained by demographic effects.

```{r tables, results = "asis"}
covar <- readRDS("~/Downloads/GTEx_v7_brain_covar.rds")
all.tables(brain05, covar)
```

## Varying the threshold

Below are plots of loadings with the nonmissingness threshold set at .05 (left) and .001 (right).

```{r compareThresh}
all.plots.comparison(brain05, brain001)
```
