---
title: "NPMLE: Some intuition"
author: "Jason Willwerscheid"
date: "5/9/2020"
output:
  workflowr::wflow_html:
    code_folding: show
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>", warning = TRUE)
```

I want to better understand the NPMLE problem. Say I have five observations, which I'll just simulate from a standard normal distribution:
$$ x_i \sim N(0, 1 - \sigma^2) + N(0, \sigma^2), $$
In this setup, $N(0, 1 - \sigma^2)$ represents the signal, which is unknown, while $N(0, \sigma^2)$ represents the noise, which is of known magnitude. There are two limit cases. In the noiseless case ($\sigma^2 = 0$), the NPMLE is just the empirical distribution. When $\sigma^2 = 1$, I'd expect to get a point mass at the sample mean (but this will of course depend on the draw and, in particular, whether the MLE estimate of the variance is greater than 1). Varying $\sigma^2$ should produce a series of distributions that are intermediate between the two. I check this. The points at the bottom of each plot mark the observations; the x marks the sample mean.

```{r ex1}
suppressMessages(library(tidyverse))

set.seed(90210)
samp <- rnorm(5)
cat("MLE estimate of variance:", mean((samp - mean(samp))^2))

cdf_grid <- seq(-2.5, 2.5, by = 0.05)
cdf_df <- tibble()

g_init <- ashr::unimix(pi = rep(.01, 100),
                       a = seq(-2.5, 2.45, by = 0.05),
                       b = seq(-2.45, 2.5, by = 0.05))

for (s2 in c(.01, .05, .1, .25, .5, .75, .9, .95, .99)) {
  ebnm_res <- ebnm::ebnm_npmle(x = samp, s = sqrt(s2), g_init = g_init)
  new_data <- tibble(x = cdf_grid, 
                     y = drop(ashr::mixcdf(ebnm_res$fitted_g, cdf_grid)), 
                     s2 = s2)
  cdf_df <- cdf_df %>% bind_rows(new_data)
}

ggplot(cdf_df, aes(x = x, y = y)) + geom_line() + 
  geom_point(data = tibble(x = samp, y = 0), shape = 1) +
  geom_point(data = tibble(x = mean(samp), y = 0), shape = 4) +
  facet_wrap(~s2)
```

I noted above that you don't always get a point mass when $\sigma^2 \approx 1$. I illustrate using a different seed.

```{r ex2}
set.seed(666)
samp <- rnorm(5)
cat("MLE estimate of variance:", mean((samp - mean(samp))^2))

cdf_df <- tibble()

for (s2 in c(.01, .05, .1, .25, .5, .75, .9, .95, .99)) {
  ebnm_res <- ebnm::ebnm_npmle(x = samp, s = sqrt(s2), g_init = g_init)
  new_data <- tibble(x = cdf_grid, 
                     y = drop(ashr::mixcdf(ebnm_res$fitted_g, cdf_grid)), 
                     s2 = s2)
  cdf_df <- cdf_df %>% bind_rows(new_data)
}

ggplot(cdf_df, aes(x = x, y = y)) + geom_line() + 
  geom_point(data = tibble(x = samp, y = 0), shape = 1) +
  geom_point(data = tibble(x = mean(samp), y = 0), shape = 4) +
  facet_wrap(~s2)
```
