---
title: "Comparing initialization functions"
author: "Jason Willwerscheid"
date: "7/25/2018"
output:
  workflowr::wflow_html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

The default function used to initialize a new factor/loading is `udv_si`, which is a simple wrapper to `softImpute` with option `type = "als"`. We previously chose to use `type = "als"` rather than `type = "svd"` because initial results showed the former to be much faster. Here I investigate further, using a large GTEx dataset as an example.

## Problem with type = "als"

I discovered an issue when investigating occasional large decreases in the FLASH objective. The dataset I use here is taken from MASH's analysis of GTEx data. `flash_add_greedy` successfully adds 19 factors, then adds  a 20th factor that improves the objective until a bad update occurs. This update increases the objective to such an extent that `nullcheck` erroneously removes the factor and FLASH terminates prematurely.

The last few lines of verbose output are as follows:
```{}
(...)
Objective:-1258100.51191229
Objective:-1258097.28313664
Objective:-1258093.37994353
Objective:-1258088.30966224
Objective:-1258081.1984392
Objective:-1258219.32382352
An iteration decreased the objective. This happens occasionally, perhaps due to numeric reasons. You could ignore this warning, but you might like to check out https://github.com/stephenslab/flashr/issues/26 for more details.performing nullcheck
objective from deleting factor:-1258152.21114343
objective from keeping factor:-1258219.32382352
factor zeroed out
```

But if I add 19 factors, then add the 20th separately, I get a much different result:
```{}
(...)
Objective:-1257665.05089993
Objective:-1257664.93935874
Objective:-1257664.72526429
Objective:-1257664.56199637
Objective:-1257664.52184646
Objective:-1257664.51611008
performing nullcheck
objective from deleting factor:-1258152.21114343
objective from keeping factor:-1257664.51611008
nullcheck complete, objective:-1257664.51611008
```

The reason for the difference is that `softImpute` randomly initializes `u` when option `type = "als"` is used. On the second try, I simply got luckier with the initialization.

To reproduce the above results, run the following (it will take some time):

```{r problem, eval=FALSE}
# devtools::install_github("stephenslab/flashr")
devtools::load_all("/Users/willwerscheid/GitHub/flashr")
# devtools::install_github("stephenslab/ebnm")
devtools::load_all("/Users/willwerscheid/GitHub/ebnm")

gtex <- readRDS(gzcon(url("https://github.com/stephenslab/gtexresults/blob/master/data/MatrixEQTLSumStats.Portable.Z.rds?raw=TRUE")))
strong <- t(gtex$strong.z)

# 20th factor zeroed out:
fl <- flash_add_greedy(strong, Kmax=50, verbose=TRUE)

# 20th factor successfully added:
fl2 <- flash_add_greedy(strong, Kmax=19, verbose=TRUE)
fl3 <- flash_add_greedy(strong, Kmax=1, f_init=fl2, verbose=TRUE)
```

## Questions for investigation

The above increase in the objective function points to a larger problem, which I will turn to in a [subsequent investigation](warmstart.html). Here, I want to revisit the choice of default `init_fn`. In particular, should we instead default to `udv_svd` (a simple wrapper to `svd`) when there is no missing data? Since `softImpute(type = "svd")` gives the same result as `svd`, and `softImpute(type = "als")` begins by calling `svd` on a random matrix anyway, using `udv_svd` can only speed things up (in addition to eliminating an annoying source of randomness). Next, when there is missing data, what are the differences between `"svd"` and `"als"` in terms of speed and the final objective attained? Preliminary results suggested that `"als"` would be much faster, but it would be useful to verify those results on the GTEx dataset.

## Results: no missing data

Since the 20th factor causes a problem for `udv_si`, I fit 19 factors greedily. I run it once with `udv_si_svd` and `udv_svd` and I run it twice with `udv_si` (using a different seed for the second run).

To reproduce the results in this section, run the following (make sure to load the `trackObj` branch of `flashr`). Since the fits take several minutes each, I pre-run the code and then load the results from file.

```{r nomissing, eval=FALSE}
# This block was run in advance.

# devtools::install_github("stephenslab/flashr", ref="trackObj")
devtools::load_all("/Users/willwerscheid/GitHub/flashr")

res.udv_si <- flash_add_greedy(strong, Kmax=19, verbose=TRUE)
res.udv_si_svd <- flash_add_greedy(strong, Kmax=19, init_fn="udv_si_svd", verbose=TRUE)
res.udv_svd <- flash_add_greedy(strong, Kmax=19, init_fn="udv_svd", verbose=TRUE)

# Change seed
res.udv_si666 <- flash_add_greedy(strong, Kmax=19, verbose=TRUE, seed=666)

all_res <- list(udv_si = res.udv_si,
                udv_si666 = res.udv_si666,
                udv_si_svd = res.udv_si_svd,
                udv_svd = res.udv_svd)

saveRDS(all_res, "../data/init_fn/all_res.rds")
```

```{r load_data}
# devtools::install_github("stephenslab/flashr", ref="trackObj")
devtools::load_all("/Users/willwerscheid/GitHub/flashr")

gtex <- readRDS(gzcon(url("https://github.com/stephenslab/gtexresults/blob/master/data/MatrixEQTLSumStats.Portable.Z.rds?raw=TRUE")))
strong <- t(gtex$strong.z)

all_res <- readRDS("./data/init_fn/all_res.rds")
```

### Initialization time

The total time (in seconds) spent initializing factors is as follows.

```{r nomissing_init_time}
init_t <- lapply(all_res, function(res) {unlist(res$init_t)})
lapply(init_t, sum)
```

As expected, `udv_svd` is fastest, but there is only about 10 seconds difference between `udv_svd` and `udv_si` (the slowest method). This difference is not hugely important given that the overall fit takes several minutes.

The initialization time per factor is as follows. 

```{r nomissing_time}
K <- length(init_t[[1]])
data <- data.frame(t = unlist(init_t), 
                   init_fn = rep(names(init_t), each=K))
data$init_fn[data$init_fn == "udv_si666"] <- "udv_si"
data$init_fn <- factor(data$init_fn)

boxplot(t ~ init_fn, data, ylim=c(0, 2), 
        main="Initialization time (s)")
```

### Number of iterations

Since the optimization step takes much more time than the initialization step, I check to see whether any initialization method is "better" in the sense that it requires less iterations to optimize. (I suppress results for `udv_si_svd` because they are identical to `udv_svd`.) 

```{r nomissing_niter}
niter <- lapply(all_res, 
                function(res) {
                  sapply(res$obj, function(i) {length(i$after_tau)})
                })
niter$udv_si_svd <- NULL

K <- length(niter[[1]])
data <- data.frame(niter = unlist(niter), 
                   init_fn = rep(names(niter), each=K))
data$init_fn[data$init_fn == "udv_si666"] <- "udv_si"
data$init_fn <- factor(data$init_fn)

boxplot(niter ~ init_fn, data, ylim=c(0, 120),
        main="Number of iterations per factor")
```
These two plots look very similar to me.

### Objective attained

There is some variation in the final objective. `softImpute(type = "als")` can do better than `svd`, but it can also do worse.

```{r nomissing_finalobj}
lapply(all_res, function(res) {flash_get_objective(strong, res$f)})
```

The difference in objective function after adding each factor/loading pair is as follows. (Again, I suppress results for `udv_si_svd` because they are identical to `udv_svd`.) 

```{r nomissing_objeach}
final.obj <- lapply(all_res, 
                    function(res) {
                      sapply(res$obj, function(obj) {
                        max(unlist(obj))
                      })
                    })
si.diff <- final.obj$udv_si - final.obj$udv_svd
si666.diff <- final.obj$udv_si666 - final.obj$udv_svd
plot(si.diff, type='l', col="darkblue",
     xlab="Factor/loading index",
     ylab="Diff. in obj. from udv_svd",
     main="Diff. in obj. using udv_si rather than udv_svd")
lines(si666.diff, col="lightblue")
legend("topright", c("seed = 123", "seed = 666"), 
       lty=c(1, 1), col=c("darkblue", "lightblue"))
abline(0, 0, lty=2)
```

Note that `udv_svd` is doing much worse at factor/loading 6, but catches up at factor/loading 7. In fact, both methods are actually adding the "same" factor/loading pairs here, but they do so in a different order:

```{r plot_factors}
par(mfrow=c(2, 2))
missing.tissues <- c(7, 8, 19, 20, 24, 25, 31, 34, 37)
gtex.colors <- read.table("https://github.com/stephenslab/gtexresults/blob/master/data/GTExColors.txt?raw=TRUE", 
                          sep = '\t', 
                          comment.char = '')[-missing.tissues, 2]
ls <- list(all_res$udv_si$f$EL[, 6], all_res$udv_si$f$EL[, 7],
        all_res$udv_svd$f$EL[, 6], all_res$udv_svd$f$EL[, 7])
mains <- list("udv_si loading 6", "udv_si loading 7",
           "udv_svd loading 6", "udv_svd loading 7")
for (i in 1:4) {
  barplot(ls[[i]], main=mains[[i]], las=2, cex.names = 0.4, 
          col=as.character(gtex.colors), names="")
}
```

I zoom in on the last ten factor/loading pairs: 

```{r nomissing_objeach2}
plot(10:19, si.diff[10:19], type='l', col="darkblue",
     xlab="Factor/loading index",
     ylab="Diff. in obj. from udv_svd")
lines(10:19, si666.diff[10:19], col="lightblue")
legend("bottomleft", c("seed = 123", "seed = 666"), 
       lty=c(1, 1), col=c("darkblue", "lightblue"))
abline(0, 0, lty=2)
```

While there are differences, there is no clear pattern here. It might be the case that, as above, factor/loading pairs are simply being added in a different order. I will revisit this question in a [subsequent investigation](init_fn2.html).


## Results: missing data

I delete 5% of the entries in the dataset and repeat the experiment. Since `svd` cannot handle missing data, I only compare `udv_si` and `udv_si_svd`.

```{r missing}
set.seed(1)
missing <- rbinom(length(strong), 1, prob=0.05)
strong.missing <- strong
strong.missing[missing] <- NA
fl_data <- flash_set_data(strong.missing)
```

```{r missing_fits, eval=FALSE}
# This block was run in advance.

res.missing.udv_si <- flash_add_greedy(fl_data, Kmax=19, verbose=TRUE)
res.missing.udv_si666 <- flash_add_greedy(fl_data, Kmax=19, verbose=TRUE, seed=666)
res.missing.udv_si_svd <- flash_add_greedy(fl_data, Kmax=19, init_fn="udv_si_svd", verbose=TRUE)

all_missing <- list(udv_si = res.missing.udv_si,
                    udv_si666 = res.missing.udv_si666,
                    udv_si_svd = res.missing.udv_si_svd)

saveRDS(all_missing, "../data/init_fn/all_missing.rds")
```

```{r load_missing}
all_missing <- readRDS("./data/init_fn/all_missing.rds")
```

### Initialization time

The total time (in seconds) spent initializing factors is as follows.

```{r missing_init_time}
init_t <- lapply(all_missing, function(res) {unlist(res$init_t)})
lapply(init_t, sum)
```

The time per factor is as follows.

```{r missing_time}
K <- length(init_t[[1]])
data <- data.frame(t = unlist(init_t), 
                   init_fn = rep(names(init_t), each=K))
data$init_fn[data$init_fn == "udv_si666"] <- "udv_si"
data$init_fn <- factor(data$init_fn)

boxplot(t ~ init_fn, data, ylim=c(0, 2),
        main="Initialization time (s)")
```

Interestingly, `udv_si_svd` seems to be reliably faster. This result contradicts our earlier results showing that using option `type = "als"` with `softImpute` was faster than using option `type = "svd"`. I'm not sure whether any code has changed in the meantime, or whether our preliminary investigations were simply faulty.

### Number of iterations, objective attained

With only 5% of data missing, the number of iterations per factor and the increase in objective per factor were nearly identical to the above, so I have suppressed them here.

To verify for yourself, run the following lines.

```{r missing_niter, eval=FALSE}
niter <- lapply(all_missing, 
                function(res) {
                  sapply(res$obj, function(i) {length(i$after_tau)})
                })

K <- length(niter[[1]])
data <- data.frame(niter = unlist(niter), 
                   init_fn = rep(names(niter), each=K))
data$init_fn[data$init_fn == "udv_si666"] <- "udv_si"
data$init_fn <- factor(data$init_fn)

boxplot(niter ~ init_fn, data, ylim=c(0, 120),
        main="Number of iterations per factor")
```

```{r missing_finalobj, eval=FALSE}
lapply(all_missing, function(res) {flash_get_objective(fl_data, res$f)})
```

```{r missing_objeach, eval=FALSE}
final.obj <- lapply(all_missing, 
                    function(res) {
                      sapply(res$obj, function(obj) {
                        max(unlist(obj))
                      })
                    })
si.diff <- final.obj$udv_si - final.obj$udv_si_svd
si666.diff <- final.obj$udv_si666 - final.obj$udv_si_svd
plot(si.diff, type='l', col="darkblue",
     xlab="Factor/loading index",
     ylab="Diff. in obj. from udv_si_svd",
     main="Diff. in obj. using udv_si rather than udv_si_svd")
lines(si666.diff, col="lightblue")
legend("topright", c("seed = 123", "seed = 666"), 
       lty=c(1, 1), col=c("darkblue", "lightblue"))
abline(0, 0, lty=2)
```

## Conclusions

There does not seem to be any advantage to using `softImpute` when there is no missing data. Indeed, `udv_si` is relatively slow and has the additional disadvantage that results depend upon a random initialization. For this reason, I recommend changing the default `init_fn` to `udv_svd` when there is no missing data.

More surprisingly, I was unable to verify earlier results that concluded that `udv_si` (that is, `softImpute` with option `type = "als"`) is faster than `udv_si_svd` (`softImpute` with option `type = "svd"`). On the contrary, `udv_si_svd` appears to be slightly faster, and again has the advantage that it is not random. Further, the algorithm is somewhat simpler to understand. Thus I recommend setting the default `init_fn` to `udv_si_svd` when data is missing.
