---
title: "Nonnegative FLASH example"
author: "Jason Willwerscheid"
date: "9/11/2018"
output:
  workflowr::wflow_html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

Nonnegative matrix factorization is straightforward in FLASH; we need only put a class of nonnegative priors on the factors and loadings. In ASH, the `+uniform` priors constitute such a class (and, at present, this is the only such class in ASH).

As an example of nonnegative matrix factorization via FLASH, I analyze the GTEx donation matrix. This matrix consists of 44 rows, each of which corresponds to a tissue or cell type (liver, lung, cortex, whole blood, etc.), and 571 columns, each of which corresponds to a donor. Note that a given donor can contribute multiple samples of a given tissue. 

In this analysis, I ignore the fact that "errors" are not, as FLASH assumes, Gaussian. I will address this simplification in subsequent analyses (see [here](count_data.html) and [here](binary_data.html)).


## Data

First I set up the donation matrix. As in previous analyses, I use GTEx v6 data rather than the more recent v7 or v8 data. 

```{r data}
raw <- read.csv("https://storage.googleapis.com/gtex_analysis_v6/annotations/GTEx_Data_V6_Annotations_SampleAttributesDS.txt",
                header=TRUE, sep='\t')

data <- raw[, c("SAMPID", "SMTSD")] # sample ID, tissue type
# Extract donor ID:
tmp <- strsplit(as.character(data$SAMPID), "-")
data$SAMPID <- as.factor(sapply(tmp, function(x) {x[[2]]})) 
names(data) <- c("DonorID", "TissueType")

data <- suppressMessages(reshape2::acast(data, TissueType ~ DonorID))

missing.tissues <- c(1, 8, 9, 20, 21, 24, 26, 27, 33, 36, 39)
data <- data[-missing.tissues, ]

gtex.colors <- read.table("https://github.com/stephenslab/gtexresults/blob/master/data/GTExColors.txt?raw=TRUE",
                          sep = '\t', comment.char = '')
gtex.colors <- gtex.colors[-c(7, 8, 19, 20, 24, 25, 31, 34, 37), 2]
gtex.colors <- as.character(gtex.colors)
```

As a preliminary step, it is useful to visualize the correlation matrix. I use Kushal Dey's `CorShrink` package, which outputs the empirical correlation matrix and shrunken estimates of the "true" correlations.

```{r corr}
tmp <- capture.output(
  CorShrink::CorShrinkData(t(data), sd_boot = TRUE, image = "both",
                           image.control = list(tl.cex = 0.25))
)
```


## Initialization function

To obtain nonnegative factor/loading pairs using FLASH, we can likely do better than to use the default SVD-type initialization. Here I use the nonnegative matrix factorization implemented in package `NNLM`.

```{r init}
udv_nn = function(Y, K = 1) {
  tmp = NNLM::nnmf(Y, K, verbose = FALSE)
  return(list(d = rep(1, K), u = tmp$W, v = t(tmp$H)))
}
```


## Greedy loadings

One round of greedily adding factors and then backfitting produces five factor/loading pairs, but repeated rounds yield additional factor/loadings. Here I do three rounds of fitting. I obtain the following loadings:

```{r greedy}
devtools::load_all("~/GitHub/flashr")
devtools::load_all("~/GitHub/ebnm")

ebnm_fn = "ebnm_ash"
ebnm_param = list(mixcompdist = "+uniform", warmstart = TRUE)

run_flash_once <- function(f_init) {
  flash(data, f_init = f_init, 
        ebnm_fn = ebnm_fn, ebnm_param = ebnm_param,
        var_type="constant", init_fn = udv_nn, 
        backfit = TRUE, verbose = FALSE)
}

fl_g <- run_flash_once(f_init = NULL)
fl_g <- run_flash_once(f_init = fl_g)
fl_g <- run_flash_once(f_init = fl_g)
fl_g$objective

plot(fl_g, plot_loadings = TRUE, loading_colors = gtex.colors,
     loading_legend_size = 3, 
     plot_grid_nrow = 4, plot_grid_ncol = 3)
```


## Backfitted loadings

Next I add 12 factors at once using `NNLM` and then backfit. Note that the objective obtained using this method is much better than the above (by approximately 500).

```{r backfit}
fl_nnmf <- flash_add_factors_from_data(data, 12,
                                       init_fn = udv_nn, 
                                       backfit = FALSE)

fl_b <- flash_backfit(data, fl_nnmf,
                      ebnm_fn = ebnm_fn,
                      ebnm_param = ebnm_param,
                      var_type = "constant",
                      verbose = FALSE)
fl_b$objective

plot(fl_b, plot_loadings = TRUE, loading_colors = gtex.colors,
     loading_legend_size = 3, 
     plot_grid_nrow = 4, plot_grid_ncol = 3)
```


## NNLM loadings

The backfitted loadings can be compared with the loadings that are obtained by simply running `NNLM`. Results are quite similar.

```{r nnmf}
plot(fl_nnmf, plot_loadings=TRUE, loading_colors = gtex.colors,
     loading_legend_size = 3, 
     plot_grid_nrow = 4, plot_grid_ncol = 3)
```


## Discussion

The loading that accounts for the largest PVE (backfitted loading 3) puts a large weight on whole blood and a smaller weight on fibroblasts. In effect, there are many more whole blood samples than samples from any other tissue (1822 versus, on average, 224). 

The loading with the second-largest PVE (loading 1) shares weights among brain tissues, with a smaller weight on pituitary tissues. FLASH finds a very similar loading in the GTEx summary data (loading 2 [here](https://willwerscheid.github.io/MASHvFLASH/MASHvFLASHnn.html#multi-tissue_effects), for example).

Loading 5 puts large weights on female reproductive tissues (ovary, uterus, and vagina), so it is most likely a sex-specific loading. Similarly, loading 4 puts its largest weights on testicular and prostate tissues. Both of these loadings put very similar weights on non-reproductive tissues (observe, in particular, the similarity of weights among gastro-intestinal tissues), so we should not make too much of correlations between reproductive and non-reproductive tissues. It is probably the case that the two sex-specific loadings are entangled with a third loading that is more properly weighted on gastro-intestinal tissues alone.
