---
title: "Bad loadings updates"
author: "Jason Willwerscheid"
date: "7/16/2018"
output:
  workflowr::wflow_html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

Here I look into the bad loadings updates some more. 

First I load the flash object from just before the "bad" update discussed in the [first investigation](objective3.html). Next, I alternately update the precision and the loadings (I do not update the factors). I do so for 20 iterations.

```{r iterate}
# devtools::install_github("stephenslab/flashr", ref="trackObj")
devtools::load_all("/Users/willwerscheid/GitHub/flashr")
# devtools::install_github("stephenslab/ebnm")
devtools::load_all("/Users/willwerscheid/GitHub/ebnm")

load("data/before_bad.Rdata")

init_fl <- res2$f
data <- flash_set_data(strong)
k <- 4

fl = init_fl
all_fls = list()
niters = 20
for (i in 1:niters) {
  # update precision
  R2 = flashr:::flash_get_R2(data, fl)
  fl$tau = flashr:::compute_precision(R2, data$missing,
                                      "by_column", data$S)
  # update loadings
  s2 = 1/(fl$EF2[, k] %*% t(fl$tau))
  s = sqrt(s2)
  Rk = flashr:::flash_get_Rk(data, fl, k)
  x = fl$EF[, k] %*% t(Rk * fl$tau) * s2
  ebnm_l = flashr:::ebnm_pn(x, s, list())
  KL_l = (ebnm_l$penloglik
          - flashr:::NM_posterior_e_loglik(x, s, ebnm_l$postmean,
                                           ebnm_l$postmean2))
  fl$EL[, k] = ebnm_l$postmean
  fl$EL2[, k] = ebnm_l$postmean2
  fl$gl[[k]] = ebnm_l$fitted_g
  fl$KL_l[[k]] = KL_l

  all_fls[[i]] = fl
}
```


## Convergence of objective

Interestingly, the objective function gets worse every iteration. Nonetheless, it is apparently converging to something.

```{r obj}
for (i in 1:niters) {
  message(flash_get_objective(data, all_fls[[i]])
          - flash_get_objective(data, init_fl))
}
```


## Convergence of prior

To confirm that convergence is taking place, I check the estimated prior $g_l$ after each iteration. The first list gives the values of $\pi_0$; the second, the values of $a$:

```{r gl}
for (i in 1:niters) {
  message(all_fls[[i]]$gl[[4]]$pi0)
}
for (i in 1:niters) {
  message(all_fls[[i]]$gl[[4]]$a)
}
```
