---
title: "Objective Function: ebnm_pn vs. ebnm_ash"
author: "Jason Willwerscheid"
date: "7/14/2018"
output:
  workflowr::wflow_html
---

## Introduction

Here I use `ebnm_ash` to see if I obtain similar decreases in the objective function as were obtained in the [previous investigation](objective.html).

## Fits

I use the same dataset as in the previous investigation.

```{r example}
# devtools::install_github("stephenslab/flashr", ref="trackObj")
devtools::load_all("/Users/willwerscheid/GitHub/flashr")
# devtools::install_github("stephenslab/ebnm")
devtools::load_all("/Users/willwerscheid/GitHub/ebnm")

gtex <- readRDS(gzcon(url("https://github.com/stephenslab/gtexresults/blob/master/data/MatrixEQTLSumStats.Portable.Z.rds?raw=TRUE")))
strong <- gtex$strong.z
```

I fit four factors greedily using both `ebnm_pn` and `ebnm_ash`.

```{r flash}
pn_res <- flash_add_greedy(strong, Kmax=4, verbose=FALSE)
ash_res <- flash_add_greedy(strong, Kmax=4, ebnm_fn = "ebnm_ash", 
                            verbose=FALSE)

plot_obj <- function(res, k, niters) {
  obj_data <- as.vector(rbind(res$obj[[k]]$after_tau,
                              res$obj[[k]]$after_f,
                              res$obj[[k]]$after_l))
  max_obj <- max(obj_data)
  obj_data <- obj_data - max_obj
  iter <- 1:length(obj_data) / 3
  
  if (length(obj_data) > niters*3) {
    idx <- (length(obj_data) - niters*3 + 1):length(obj_data)
    obj_data <- obj_data[idx]
    iter <- iter[idx]
  }
  
  plt_xlab <- "Iteration"
  plt_ylab <- "Diff. from maximum obj."
  plt_colors <- c("indianred1", "indianred3", "indianred4")
  plt_pch <- c(16, 17, 15)
  main <- paste("Factor/loading", k)
  
  plot(iter, obj_data, col=plt_colors, pch=plt_pch,
       xlab=plt_xlab, ylab=plt_ylab, main=main)
  legend("bottomright", c("after tau", "after f", "after l"),
         col=plt_colors, pch=plt_pch)
}
```

## Results: ebnm_pn

The problem discussed in the [previous investigation](objective.html) occurs every time.

```{r plot_pn}
plot_obj(pn_res, 1, niters=3)
plot_obj(pn_res, 2, niters=5)
plot_obj(pn_res, 3, niters=20)
plot_obj(pn_res, 4, niters=10)
```

## Results: ebnm_ash

But no obvious problems occur when using `ebnm_ash`.

```{r plot_ash}
plot_obj(ash_res, 1, niters=3)
plot_obj(ash_res, 2, niters=5)
plot_obj(ash_res, 3, niters=10)
plot_obj(ash_res, 4, niters=20)
```

## Conclusions

When using `ebnm_ash`, the objective does not suffer from the same erratic behavior as when using `ebnm_pn`. Is there a weird bug somewhere in the computation of the likelihood function for `ebnm_pn`?
