---
title: "Storing tau as a scalar"
author: "Jason Willwerscheid"
date: "9/16/2018"
output:
  workflowr::wflow_html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In `flashr` [Issue #83](https://github.com/stephenslab/flashr/issues/83), I identify some possible ways to speed up fitting when `var_type = "constant"` or when `var_type = "zero"` and `S` (the standard errors) is stored as a scalar. I also identify ways to reduce memory usage when loadings or factors are not fixed (and they will almost never both be fixed).

Here I implement the suggested changes and profile the code on data from the GTEx project. I compare elapsed time and total memory used in five scenarios: 1. `var_type = "by_column"`; 2. `var_type = "constant"`; 3. `var_type = "zero"` (with standard errors fixed); 4. `var_type = "constant"` with missing data; and 5. `var_type = "by_column"` with fixed factors. (The fifth scenario was added to help ensure that no bugs were introduced into the code; little improvement is expected.)

## Details

First I load the data and generate missing entries.

```{r data}
devtools::load_all("~/GitHub/ebnm")
library("profmem")

gtex <- readRDS(gzcon(url("https://github.com/stephenslab/gtexresults/blob/master/data/MatrixEQTLSumStats.Portable.Z.rds?raw=TRUE")))

set.seed(666)

strong <- gtex$strong.z
strong_missing <- strong
strong_missing[sample(1:length(strong), 
                    floor(0.1*length(strong)), 
                    replace=FALSE)] <- NA

K <- 3 # number of factors to greedily add in scenarios 1-4
FF <- cbind(rep(1, 44),
            c(rep(0, 6), rep(NA, 10), rep(0, 28)),
            c(rep(0, 43), 1)) # fixed factors for scenario 5

ntrials <- 5
```

Next I run the code from the `master` branch. In each case, I add three factors greedily and then backfit.

```{r before}
system("cd ~/GitHub/flashr; git checkout master")
devtools::load_all("~/GitHub/flashr")

data <- flash_set_data(strong)
data_fixS <- flash_set_data(strong, S = 1)
data_missing <- flash_set_data(strong_missing)

t_bycol_bef <- system.time({
  mem_bycol_bef <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_bycol_before <- flash(data, Kmax=K,
                               var_type="by_column",
                               backfit=TRUE, verbose=FALSE)
    }
  })
})

t_const_bef <- system.time({
  mem_const_bef <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_const_before <- flash(data, Kmax=K,
                               var_type="constant",
                               backfit=TRUE, verbose=FALSE)
    }
  })
})

t_zero_bef <- system.time({
  mem_zero_bef <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_zero_before <- flash(data_fixS, Kmax=K, 
                              var_type="zero",
                              backfit=TRUE, verbose=FALSE)
    }
  })
})

t_miss_bef <- system.time({
  mem_miss_bef <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_miss_before <- flash(data_missing, Kmax=K, 
                              var_type="constant",
                              backfit=TRUE, verbose=FALSE)
    }
  })
})

t_fixed_bef <- system.time({
  mem_fixed_bef <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_fixed_before <- flash_add_fixed_factors(data_missing, FF, 
                                                 backfit=TRUE,
                                                 verbose=FALSE)
    }
  })
})
```

Finally, I load the branch with the proposed changes and re-run the code.

```{r after}
system("cd ~/GitHub/flashr; git checkout efficient-tau2")
devtools::load_all("~/GitHub/flashr")

data <- flash_set_data(strong)
data_fixS <- flash_set_data(strong, S = 1)
data_missing <- flash_set_data(strong_missing)

t_bycol_aft <- system.time({
  mem_bycol_aft <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_bycol_after <- flash(data, Kmax=K,
                              var_type="by_column",
                              backfit=TRUE, verbose=FALSE)
    }
  })
})

t_const_aft <- system.time({
  mem_const_aft <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_const_after <- flash(data, Kmax=K,
                              var_type="constant",
                              backfit=TRUE, verbose=FALSE)
    }
  })
})

t_zero_aft <- system.time({
  mem_zero_aft <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_zero_after <- flash(data_fixS, Kmax=K, 
                             var_type="zero",
                             backfit=TRUE, verbose=FALSE)
    }
  })
})

t_miss_aft <- system.time({
  mem_miss_aft <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_miss_after <- flash(data_missing, Kmax=K, 
                             var_type="constant",
                             backfit=TRUE, verbose=FALSE)
    }
  })
})

t_fixed_aft <- system.time({
  mem_fixed_aft <- profmem::profmem({
    for (i in 1:ntrials) {
      fl_fixed_after <- flash_add_fixed_factors(data_missing, FF,
                                                backfit=TRUE,
                                                verbose=FALSE)
    }
  })
})
```

## Results

The elapsed time per trial (in seconds) is as follows.

```{r time}
rnames <- c("before", "after")
cnames <- c("by_column", "constant", "zero", "missing", "fixed")
t_table <- t(matrix(c(t_bycol_bef[3], 
                      t_const_bef[3], 
                      t_zero_bef[3], 
                      t_miss_bef[3],
                      t_fixed_bef[3],
                      t_bycol_aft[3], 
                      t_const_aft[3],
                      t_zero_aft[3], 
                      t_miss_aft[3],
                      t_fixed_aft[3]),
                    nrow=5, ncol=2)) / ntrials
rownames(t_table) <- rnames
colnames(t_table) <- cnames
round(t_table, digits = 1)
```

The memory used per trial (in GB) is as follows.

```{r mem}
mem_table <- t(matrix(c(sum(mem_bycol_bef$bytes, na.rm=TRUE),
                        sum(mem_const_bef$bytes, na.rm=TRUE),
                        sum(mem_zero_bef$bytes, na.rm=TRUE),
                        sum(mem_miss_bef$bytes, na.rm=TRUE),
                        sum(mem_fixed_bef$bytes, na.rm=TRUE),
                        sum(mem_bycol_aft$bytes, na.rm=TRUE),
                        sum(mem_const_aft$bytes, na.rm=TRUE),
                        sum(mem_zero_aft$bytes, na.rm=TRUE),
                        sum(mem_miss_aft$bytes, na.rm=TRUE),
                        sum(mem_fixed_aft$bytes, na.rm=TRUE)),
                      nrow=5, ncol=2)) / 1024^3 / ntrials
rownames(mem_table) <- rnames
colnames(mem_table) <- cnames
round(mem_table, digits = 1)
```

Finally, I run a quick check to verify that I am getting the same flash object in each case.

```{r check}
identical(fl_bycol_before$objective, fl_bycol_after$objective)
identical(fl_const_before$objective, fl_const_after$objective)
identical(fl_zero_before$objective, fl_zero_after$objective)
identical(fl_miss_before$objective, fl_miss_after$objective)
identical(fl_fixed_before$objective, fl_fixed_after$objective)
```

## Discussion

As expected, I obtain substantial speed-ups when `var_type = "constant"` and when `var_type = "zero"`. (In fact, there is an improvement in each case, even when loadings are fixed).

Also as expected, memory usage is substantially reduced in all cases. 
