---
title: "Accelerating backfits with SQUAREM and DAAREM"
author: "Jason Willwerscheid"
date: "8/17/2018"
output: 
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis is complementary to my investigations into parallelization (see [here](parallel.html) and [here](parallel2.html)) in that I further explore ways to speed up FLASH backfits.

I use two off-the-shelf EM (and MM) accelerators, SQUAREM and DAAREM. For SQUAREM details, see [Varadhan and Roland (2008)](https://www.jstor.org/stable/41548597). For DAAREM, see [Henderson and Varadhan (2018)](https://arxiv.org/abs/1803.06673).

## Experiments

I use the same GTEx dataset that I use [here](https://willwerscheid.github.io/MASHvFLASH/MASHvFLASHgtex.html) and in [Investigation 8](parallel.html).

I use `flash_add_greedy` to create three flash fit objects with, respectively, 5, 10, and 20 factor/loading pairs. I then refine each fit using `flash_backfit` with: 1. no acceleration; 2. acceleration via SQUAREM; and 3. acceleration via DAAREM. 

## Results

Since the experiments take a long time to run, I pre-run the code [below](#code) and load the results from file.

```{r load_all}
res5 <- readRDS("./data/squarem/res5.rds")
res10 <- readRDS("./data/squarem/res10.rds")
res20 <- readRDS("./data/squarem/res20.rds")
```

### Convergence behavior

In every case, DAAREM takes the fewest iterations to converge and convergence is nearly monotonic. In constrast, SQUAREM evinces highly non-monotonic behavior and tends to take more iterations than backfitting with no acceleration at all.

```{r conv}
plot_obj <- function(res, main) {
  data <- c(res$backfit_obj, res$squarem_obj$V1, res$daarem_obj)
  plot(1:length(res$backfit_obj), res$backfit_obj,
       type='l', col='red', ylim=c(min(data), max(data)),
       xlab="Iteration", ylab="Objective", main=main)
  lines(1:length(res$squarem_obj$V1), res$squarem_obj$V1,
        col='blue')
  lines(1:length(res$daarem_obj), res$daarem_obj,
        col='green')
  legend("bottomright", legend=c("DAAREM", "No acceleration", "SQUAREM"),
         lty=1, col=c('green', 'red', 'blue'))
}

plot_obj(res5, "5 factor model")
plot_obj(res10, "10 factor model")
plot_obj(res20, "20 factor model")
```

### Final objective 

Since the final objectives attained are difficult to see in the plots above, I list the differences in the table below. In every case, DAAREM beats the final objective attained with no acceleration. SQUAREM does much worse on the 5- and 10-factor models, but better on the 20-factor model.

```{r final_obj}
final_obj <- function(res) {
  backfit = res$backfit_obj[length(res$backfit_obj)]
  return(c(res$daarem_obj[length(res$daarem_obj)] - backfit,
           res$squarem_obj$V1[length(res$squarem_obj$V1)] - backfit))
}
table_data <- cbind("5_factors" = final_obj(res5), 
                    "10_factors" = final_obj(res10), 
                    "20_factors" = final_obj(res20))
rownames(table_data) = c("DAAREM (diff from backfit)", 
                         "SQUAREM (diff from backfit)")
knitr::kable(table_data, digits=1)
```

### Time per iteration

The acceleration methods add a bit of overhead in terms of time required per iteration, but not much. The following are seconds required per iteration:

```{r t}
t <- function(res) {
  return(c(res$backfit_t[3] / length(res$backfit_obj),
           res$daarem_t[3] / length(res$daarem_obj),
           res$squarem_t[3] / length(res$squarem_obj$V1)))
}
table_data <- cbind("5_factors" = t(res5), 
                    "10_factors" = t(res10), 
                    "20_factors" = t(res20))
rownames(table_data) <- c("backfit", "DAAREM", "SQUAREM")
knitr::kable(table_data, digits = 2)
```

### Total iterations to convergence

Perhaps the most important consideration is how quickly we can obtain an estimate that is reasonable close to convergence, where "reasonably close" is defined by some stopping rule. Below I give the number of iterations required given different tolerance parameters (i.e., the number of iterations required before the difference in objective from one iteration to the next is less than `tol`). (I omit results for SQUAREM since it does so poorly in other respects.)

```{r niter}
niter <- function(res, tols) {
  backfit_diff <- (res$backfit_obj[2:length(res$backfit_obj)] -
                     res$backfit_obj[1:(length(res$backfit_obj) - 1)])
  daarem_diff <- (res$daarem_obj[2:length(res$daarem_obj)] -
                     res$daarem_obj[1:(length(res$daarem_obj) - 1)])
  res <- matrix(NA, nrow = length(tols), ncol = 2)
  for (i in 1:length(tols)) {
    res[i, 1] <- min(which(backfit_diff < tols[i])) + 1
    res[i, 2] <- min(which(abs(daarem_diff) < tols[i])) + 1
  }
  rownames(res) = paste("tol =", as.character(tols))
  colnames(res) = c("backfit", "DAAREM")
  return(res)
}

table_data <- niter(res5, c(0.5, 0.1, 0.05))
knitr::kable(table_data, caption = "5 factors")

table_data <- niter(res10, c(0.5, 0.1, 0.05))
knitr::kable(table_data, caption = "10 factors")

table_data <- niter(res20, c(0.5, 0.1, 0.05))
knitr::kable(table_data, caption = "20 factors")
```

## Code

Click "Code" to view the code used to produce the above results.

```{r code, code=readLines("../code/squarem.R"), eval=FALSE}
```
