---
title: "Using warmstarts with ebnm_ash"
author: "Jason Willwerscheid"
date: "7/26/2018"
output:
  workflowr::wflow_html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

I [previously](warmstart.html) concluded that using warmstarts is more advantageous than not when `ebnm_fn = ebnm_pn`. Here I repeat the experiment with `ebnm_fn = ebnm_ash`.

## Results

```{r load_data}
# devtools::install_github("stephenslab/flashr", ref="trackObj")
devtools::load_all("/Users/willwerscheid/GitHub/flashr")
# devtools::install_github("stephenslab/ebnm")
devtools::load_all("/Users/willwerscheid/GitHub/ebnm")

gtex <- readRDS(gzcon(url("https://github.com/stephenslab/gtexresults/blob/master/data/MatrixEQTLSumStats.Portable.Z.rds?raw=TRUE")))
strong <- t(gtex$strong.z)
```

```{r fits, eval=FALSE}
# This block was run in advance.

res.no.warmstart <- flash_add_greedy(strong, Kmax=50, init_fn="udv_svd",
                                     ebnm_fn="ebnm_ash", verbose=TRUE)
res.warmstart <- flash_add_greedy(strong, Kmax=50, init_fn="udv_svd",
                                  ebnm_fn="ebnm_ash", warmstart=TRUE, 
                                  verbose=TRUE)

saveRDS(res.no.warmstart, "../data/warmstart2/nowarmstart.rds")
saveRDS(res.warmstart, "../data/warmstart2/warmstart.rds")
```

```{r load_results}
res.no.warmstart <- readRDS("./data/warmstart2/nowarmstart.rds")
res.warmstart <- readRDS("./data/warmstart2/warmstart.rds")
```

### Optimization time

The total time (in seconds) needed to optimize factors is:

```{r total_time}
x1 <- unlist(res.no.warmstart$opt_time)
x2 <- unlist(res.warmstart$opt_time)
list(no.warmstart = sum(x1), warmstart = sum(x2))
```

The time required per factor/loading is as follows.

```{r time_per_factor}
plot(x1, ylim=c(0, max(x1) + 1), pch=19, col="blue",
     xlab="Factor/loading index", ylab="Optimization time (s)")
points(x2, pch=17, col="red")
legend("topright", c("No warmstart", "Warmstart"),
       pch=c(19, 17), col=c("blue", "red"))
```

So using a warmstart often provides a speed-up, but not nearly as reliably as with `ebnm_pn`. Indeed, for several factor/loading pairs (17, 19, 22, 24, 26), the warmstart considerably slows things down.

### Objective attained

As with `ebnm_pn`, using warmstarts yields a slightly worse overall objective. 

```{r obj}
list(no.warmstart = flash_get_objective(strong, res.no.warmstart$f),
     warmstart = flash_get_objective(strong, res.warmstart$f))
```

The per-factor difference in objective is as follows.

```{r obj_diff}
o1 <- sapply(res.no.warmstart$obj, 
             function(obj) {max(unlist(obj))})
o2 <- sapply(res.warmstart$obj, 
             function(obj) {max(unlist(obj))})

plot(o2 - o1, type='l',
     xlab="Factor/loading index",
     ylab="Diff. in obj. using warmstart",
     main="Difference in overall objective after adding each factor")
```

## Conclusions

The advantages of using warmstarts are not nearly as compelling as they were with `ebnm_pn`. The speed-up is not as reliable, and the sacrifice in the final objective attained is greater. Further, I have not seen `Rmosek` fail for `ebnm_ash` in the same way that `optim` fails for `ebnm_pn`. I am not convinced that we should automatically use warmstarts for every iteration when `ebnm_fn = ebnm_ash`. Perhaps we could add a parameter `use_warmstarts` to the main FLASH functions.
